Thread Management for High Performance Database Systems - Design and Implementation \\{\large Technical Report

### 2018-08-17 Presentations/Paper/paper_thread_pool.tex
Feedback implemented as given by Gunter

Parallel data processing can be achieved by different approaches, like instruction and data parallelism or multi threading. In this paper, we focus on multi threading by implementing a thread pool for the graph database system GeckoDB. The thread pool will be integrated into BOLSTER, a high performance library for parallel execution of primitives like for or filter on large data sets. In the current implementation, BOLSTER creates a fix number of threads for each call of a primitive. This approach is called \emph{thread-per-request}. Since many primitives are executed at the same time, a couple of drawbacks arise from this implementation.
First of all, the creation of threads comes along with overhead like stack initialisation and memory allocation. Secondly, creating a huge number of threads simultaneously may lead to large context switch overhead of the scheduler. Additionally, debugging and profiling applications that create many threads during runtime is 	time-consuming.

@@

In this work we make the following contributions:
\begin{itemize}
	\item \textbf{Design and Implementation} We describe our design and implementation of the thread pool
	\item \textbf{Waiting Strategies} We evaluate the possibility to wait for a group of task in the calling thread
	\item \textbf{Evaluation} We compare our thread pool against the existing implementation in BOLSTER
	\item \textbf{Statistics per Configuration} We measure and evaluate additional statistics for different thread pool configurations
\end{itemize}
We organized the rest of the paper as follows. In Section 2, we give preliminaries about the considered task configuration and about thread safe access of memory. In Section 3, we show our design and implementation of the thread pool and examine our experimental environment in Section 4. In Section 5, we describe the results of our performance evaluation. In Section 6, we name related work and state our conclusion and future work in Section 7.

\section{Preliminaries}
In this section, we define our configuration of tasks that are processed by the thread pool and state difficulties of synchronizing thread access to memory.

@@

\subsection{Managing Thread States}
As mentioned before, accessing the same memory location from different threads may lead to race conditions, which means the program behaviour depends on the accessing order of the threads. For example, while one thread updates data at a specific memory location, another thread writes into the data, which results to inconsistency. In Section 2, we presented two approaches to handle multithreaded access to the same memory location, mutex threads and atomic operations and we state where those approaches are used in our implementation. In this subsection, we focus on atomic operations to manage thread states. We consider the following steps for a thread to update its state. At first, the current state has to be checked and, depending on the current state, it has to be updated. If these steps are implemented using an if statement and an update function, the complete procedure is not atomic. Consequently, another thread can update the state between check of the current state and update. To avoid this, we use the function \emph{atomic\_compare\_exchange\_strong} \cite{atomicstrong} of the stdatomic library. This function combines the check and update in one atomic procedure. We implemented every thread state access with this function to ensure atomic thread state updates.

@@

\subsection{Performance monitoring}
The performance monitoring component of the thread pool system is responsible for collecting data about threads, tasks, and the thread pool instance. We will use the information collected in this component for analysis and performance optimization in later stages. As mentioned before, the performance monitoring is an optional component, since evaluating the following statistics can reduce the execution performance of our thread pool. We divide the collected statistics into the sections \emph{task statistics}, \emph{thread statistics}, and \emph{thread pool statistics}.

@@

\subsection{Testing and Memory Check}
As last aspect of our implementation, we show how we test the thread pool and how we ensure to avoid memory leaks. Since GeckoDB already uses GTest \cite{sen2010quick} for unit testing, we also wrote our test using this framework. GTest provides a simple, powerful library to perform unit tests in C++. We wrote unit tests for the priority queue, the thread pool itself and the performance monitoring. As mentioned before, we use functions from the stdatomic library of c. Using this library in GTest results in compiler errors with GCC, since the c atomic implementation differs from the c++ one of the standard library and the GCC developers will not fix this issue. Consequently, we wrote a workaround not using the c stdatomic library in the GTest compilation unit.

Since c as low level language does not have an own memory management, memory leaks may occur using heap space allocations. To find and remove all memory leaks in our implementation, we use the tool \emph{valgrind}, which offers different functionalities to find race and jump condition and to count unfreed heap blocks. We found memory leaks in our priority queue tests and also in the deallocation of thread pool instances. In the following sections, we show our experimental setup and the results of our evaluation.

@@

Additionally, we evaluate the metrics for threads, tasks and the thread pool mentioned in Section 3.F regarding the thread pool and task configuration used in the BOLSTER primitives. To evaluate optimal configurations for the thread pool, we consider the relation between waiting and completion time of  thread pool instances with different sizes.

@@

\section{Evaluation}
In this Section, we present the results of our evaluation according to the setup we stated in the previous section. We start with an introduction of the benchmark methods to compare our thread pool against the baseline implementation and present the results. Furthermore, we show the relation of waiting time and completing regarding different thread pool configurations.

@@

\section{Related Work}
Implementing multi threading can be realised using different approaches from spawning new threads to reuse them in thread pools. Pyarali et al. show that a thread pool model can significantly improve system performance and reduce response time \cite{pyarali2001evaluating}. They present different approaches to implement thread pools and show the dependency on use cases of these implementations.

@@

\section{Conclusion and Future Work}
In this work, we describe our implementation of a thread pool system. We present in detail the implementation and show our evaluation results containing a comparison against the baseline implementation and the waiting time behaviour over different thread pool sizes. Within our implementation, we state the differences of parallel memory access as well as the problem of implementing synchronous waiting for a set of tasks.

>>>>

Parallel data processing can be achieved by different approaches, like instruction and data parallelism or multi threading. In this paper, we focus on multi threading by implementing a thread pool for the graph database system GeckoDB\footnote{GeckoDB source code repository: \url{https://github.com/geckodb}}. The thread pool will be integrated into BOLSTER, a high performance library for parallel execution of primitives like for or filter on large data sets. BOLSTER has similarities to Intel TBB~\cite{books/daglib/0018624} but is handwritten from scratch to fit perfectly into GeckoDBs storage engine written in C11~\cite{pinneckegridtables}. In the current implementation, BOLSTER creates a fix number of threads for each call of a primitive. This approach is called \emph{thread-per-request}. Since many primitives are executed at the same time, a couple of drawbacks arise from this implementation.
First of all, the creation of threads comes along with overhead like stack initialisation and memory allocation. Secondly, creating a huge number of threads simultaneously may lead to large context switch overhead of the scheduler. Additionally, debugging and profiling applications that create many threads during runtime is 	time-consuming.

@@

In this work we make the following contributions:
\begin{itemize}
	\item \textbf{Design and Implementation} We describe our design and implementation of the thread pool
	\item \textbf{Waiting Strategies} We evaluate the possibility to wait for a group of task in the calling thread
	\item \textbf{Evaluation} We compare our thread pool against the existing implementation in BOLSTER
	\item \textbf{Statistics per Configuration} We measure and evaluate additional statistics for different thread pool configurations
\end{itemize}
We organized the rest of the paper as follows. In Section \ref{sec:pre}, we give preliminaries about the considered task configuration and about thread safe access of memory. In Section III, we show our design and implementation of the thread pool and examine our experimental environment in Section IV. In Section \ref{sec:eval}, we describe the results of our performance evaluation. In Section \ref{sec:related}, we name related work and state our conclusion and future work in Section \ref{sec:future}.

\section{Preliminaries}\label{sec:pre}
In this section, we define our configuration of tasks that are processed by the thread pool and state difficulties of synchronizing thread access to memory.

@@

\subsection{Managing Thread States}
As mentioned before, accessing the same memory location from different threads may lead to race conditions, which means the program behaviour depends on the accessing order of the threads. For example, while one thread updates data at a specific memory location, another thread writes into the data, which results to inconsistency. In Section \ref{sec:pre}, we presented two approaches to handle multithreaded access to the same memory location, mutex threads and atomic operations and we state where those approaches are used in our implementation. In this subsection, we focus on atomic operations to manage thread states. We consider the following steps for a thread to update its state. At first, the current state has to be checked and, depending on the current state, it has to be updated. If these steps are implemented using an if statement and an update function, the complete procedure is not atomic. Consequently, another thread can update the state between check of the current state and update. To avoid this, we use the function \emph{atomic\_compare\_exchange\_strong} \cite{atomicstrong} of the stdatomic library. This function combines the check and update in one atomic procedure. We implemented every thread state access with this function to ensure atomic thread state updates.

@@

\subsection{Performance monitoring}\label{subsec:monitor}
The performance monitoring component of the thread pool system is responsible for collecting data about threads, tasks, and the thread pool instance. We will use the information collected in this component for analysis and performance optimization in later stages. As mentioned before, the performance monitoring is an optional component, since evaluating the following statistics can reduce the execution performance of our thread pool. We divide the collected statistics into the sections \emph{task statistics}, \emph{thread statistics}, and \emph{thread pool statistics}.

@@

\subsection{Testing and Memory Check}
As last aspect of our implementation, we show how we test the thread pool and how we ensure to avoid memory leaks. Since GeckoDB already uses GTest \cite{sen2010quick} for unit testing, we also wrote our test using this framework. GTest provides a simple, powerful library to perform unit tests in C++. We wrote unit tests for the priority queue, the thread pool itself and the performance monitoring. As mentioned before, we use functions from the stdatomic library of C. Using this library in GTest results in compiler errors with GCC, since the C atomic implementation differs from the C++ one of the standard library and the GCC developers will not fix this issue. Consequently, we wrote a workaround not using the C stdatomic library in the GTest compilation unit.

Since C as low level language does not have an own memory management, memory leaks may occur using heap space allocations. To find and remove all memory leaks in our implementation, we use the tool \emph{valgrind}, which offers different functionalities to find race and jump condition and to count unfreed heap blocks. We found memory leaks in our priority queue tests and also in the deallocation of thread pool instances. In the following sections, we show our experimental setup and the results of our evaluation.

@@

Additionally, we evaluate the metrics for threads, tasks and the thread pool mentioned in Section \ref{subsec:monitor} regarding the thread pool and task configuration used in the BOLSTER primitives. To evaluate optimal configurations for the thread pool, we consider the relation between waiting and completion time of  thread pool instances with different sizes.

@@

\section{Evaluation}\label{sec:eval}
In this section, we present the results of our evaluation according to the setup we stated in the previous section. We start with an introduction of the benchmark methods to compare our thread pool against the baseline implementation and present the results. Furthermore, we show the relation of waiting time and completing regarding different thread pool configurations.

@@

\section{Related Work}\label{sec:related}
Implementing multi threading can be realised using different approaches from spawning new threads to reuse them in thread pools. Pyarali et al. show that a thread pool model can significantly improve system performance and reduce response time \cite{pyarali2001evaluating}. They present different approaches to implement thread pools and show the dependency on use cases of these implementations.

@@

\section{Conclusion and Future Work}\label{sec:future}
In this work, we describe our implementation of a thread pool system. We present in detail the implementation and show our evaluation results containing a comparison against the baseline implementation and the waiting time behaviour over different thread pool sizes. Within our implementation, we state the differences of parallel memory access as well as the problem of implementing synchronous waiting for a set of tasks.
